---
title: "Geog 456 Project"
author: "jed"
date: "1/30/2020"
output:
  html_document:
    df_print: paged
  hmtl_document: default
  md_document:
    variant: markdown_github
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##NCDOT Bike Crash Data Time Plotting

This is a tutorial for working with NCDOT bike crash data to plot change over time using RStudio.

First, we need to import the data.

```{r}
#Locate the file on your computer and name it something. We've chosen to name it "the_file"
the_file <- "C:/Users/jed/Documents/School/Senior/Spring/GEOG 456/bikecrash.csv"

#Next, open the .csv using the following line
df <- read.csv(the_file)

#head(df) will allow us to view the column names and first 6 rows of the data frame we opened
head(df)
```

Packages are groups of opensource code that can be used within R. We need to download a few before moving forward.

```{r}
#There are a number of packages we will need to make sure we have downloaded before creating our 
#visualizations. The lines below will install the packages we need and open their libraries.
install.packages("ggplot2")
install.packages("RColorBrewer")
install.packages("tidyverse")
library(ggplot2)
library(RColorBrewer)
library(tidyverse)
```

Now, let's get to the temporal analyses. We can start by choosing the Locality Type variable to measure
its change over time.


Next, we want to create a variable that counts the number of crashes that occur in each Locality
type so we are able to analyze it. The code below will create a variable called "monthly_count" 
that will do this by each month.

```{r}
monthly_count <- df %>%
  count(CrashMonth, Locality)
```

Now lets plot the data. Our data is monthly_count, and we are going to put the CrashMonth on the 
x-axis and the number of crashes (n) on the y-axis. Grouping by Locality will separate our lines
based on the Locality variable, and color = Locality will give each line a different color. The
final part of the code, geom_line(), tells R to display our data as lines.

```{r}
ggplot(data = monthly_count, mapping = aes(x = CrashMonth, y = n, group = Locality, color = Locality)) +
       geom_line()
```

This output looks good, but it isn't all that interesting. We might be able to learn more about
when people crash if we display the data by hour rather than by month. Let's create a variable
called hourly_count that does the same thing as monthly_count, but for every hour.

```{r}
hourly_count <- df %>%
  count(CrashHour, Locality)
```

Now let's plot it again, but this time using hourly_count as our data.

```{r}
ggplot(data = hourly_count, mapping = aes(x = CrashHour, y = n, group = Locality, color = Locality)) +
  geom_line()
```

Now we're getting somewhere. We can see that there are many more crashes in urbanized areas, and the 
rate of crash increase is also much higher during rush hour in urbanized areas.
Let's try to make our output easier to look at. Using the labs function, we can add a title as well as
x- and y-axis labels. The theme_bw() function changes our chart from the default style to one that is
more aesthetically pleasing. We can also change the relative size, color, and orientation of axis text
using the theme() function as outlined below.

```{r}
ggplot(data = hourly_count, mapping = aes(x = CrashHour, y = n, group = Locality, color = Locality)) +
  geom_line() +
  labs(title = "Number of Crashes by Locality Type Throughout the Day",
       x = "Hour",
       y = "Crash Frequency") +
  theme_bw() +
  theme(axis.text.x = element_text(colour = "grey20", size = 12, angle = 45, hjust = 0.5, vjust = 0.5),
        axis.text.y = element_text(colour = "grey20", size = 12),
        text = element_text(size = 16))
```

This final output is useful for analysis and tells us a story about the data and how Locality type
crash frequency changes throughout the day.
